---
title: Mitochondrial gene content annotated by MAKER
author: Shaun Jackman
output:
  html_document:
    keep_md: yes
    toc: yes
---

# Load libraries
```{r load-libraries, include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(magrittr)
library(RColorBrewer)
library(readr)
library(scales)
library(stringr)
library(tidyr)

# Bioconductor
library(genomeIntervals)
```

```{r knitr, include=FALSE}
knit_print.data.frame <- function(x, ...) kable(x) %>% paste(collapse = "\n") %>% asis_output
opts_chunk$set(tidy = FALSE, fig.width = 10, fig.height = 5)
```

# Gene families
```{r gene-families}
# Four shades each of blue, green, red, brown.
family_colour <- tribble(
	~Family_short, ~Colour,
	"atp", "#B6D4E2", "dpo", "#7EADBD", "rpo", "#39798C", "mat", "#135160",
	"rpl", "#B6E2CF", "rps", "#7EBD9C", "rrn", "#398C60", "trn", "#136034",
	"nad", "#E2B6DB", "sdh", "#BD7EB5", "cob", "#8C3984", "cox", "#60135B",
	"ccm", "#E2CCB5", "mtt", "#BD9D7E", "psa", "#8C6239", "ymf", "#603813",
	"orf", "#C0C0C0", "oth", "#808080")

gene_types <- tribble(
	~Family_short, ~Family,
	'atp', 'ATP synthase',
	'ccb', 'cytochrome c biogenesis',
	'ccm', 'cytochrome c maturation',
	'clp', 'maturases (clp)',
	'cob', 'complex III (ubichinol cytochrome c reductase)',
	'cox', 'complex IV (cytochrome c oxidase)',
	'dpo', 'DNA polymerase',
	'mat', 'maturases',
	'mtt', 'membrane targeting and translocation',
	'nad', 'complex I (NADH dehydrogenase)',
	'ori', 'origin of replication',
	'orf', 'open reading frames',
	'psa', 'photosystem I',
	'rpl', 'ribosomal proteins (LSU)',
	'rpo', 'RNA polymerase',
	'rps', 'ribosomal proteins (SSU)',
	'sdh', 'complex II (succinate dehydrogenase)',
	'ymf', 'uncharacterized mitochondrial protein',
	'oth', 'other genes',
	'rrn', 'ribosomal RNAs',
	'trn', 'transfer RNAs') %>%
	right_join(family_colour, by = "Family_short") %>%
	mutate(Family_both = paste(Family_short, Family, sep = ": "))
```

# Load data
```{r load-data}
gff_all <- readGff3("pg29mt-scaffolds.gff", isRightOpen = FALSE, quiet = TRUE)

gff <- gff_all[gff_all$type %in% c('mRNA', 'rRNA', 'tRNA'),]

scaffold_sizes <- read_tsv("pg29mt-scaffolds.fa.fai",
		col_names = c("Scaffold", "Size", "Pos", "Width", "Stride"),
		col_types = c("ciiii")) %>%
	transmute(Scaffold, ScaffoldSize = Size,
		ScaffoldNameAndSize = paste(Scaffold, " ",
			round(ScaffoldSize / 1e3), "kbp") %>% factor(., .))

data <- tibble(
	Scaffold = as.character(gff$seq_name),
	Strand = as.character(gff$strand),
	Type = as.character(gff$type),
	ID = as.character(getGffAttribute(gff, 'ID')),
	Name = as.character(getGffAttribute(gff, 'Name')),
	Position = as.integer(as.matrix(gff)[,1]),
	Size = as.integer(size(gff))) %>%
	mutate(
		Gene =  sub("[-_|].*", "", Name),
		Family_short = tolower(substr(Name, 1, 3))) %>%
	left_join(gene_types, by = "Family_short") %>%
	replace_na(list(Family = "other genes", Family_both = "oth: other genes"))
```

# Aggregate the data
```{r aggregate}
data_grouped <- data %>%
	group_by(Type, Family) %>%
	summarise(
		Count = n(),
		Size = sum(Size)) %>%
	droplevels %>%
	right_join(gene_types, ., by = "Family")
data_grouped_without_orf <- data_grouped %>% filter(Family_short != "orf")
```

# Tabulate genes by type
```{r tabulate-genes-by-type}
tab <- data %>%
	group_by(Type) %>%
	summarise(
		Count = n(),
		Size = sum(Size)) %>%
	set_rownames(.$Type) %>%
	select(Count, Size)
tab['Sum',] <- colSums(tab)
kable(tab)
```

# Tabulate genes by family
```{r tabulate-genes-by-family}
tab <- data_grouped %>%
	filter(Family_short != "orf") %>%
	set_rownames(paste(.$Family_short, .$Family)) %>%
	ungroup %>%
	select(Count, Size)
tab['Sum',] <- colSums(tab)
kable(tab)
```

# Tabulate ribosomal RNA (rRNA) genes
```{r tabulate-rrna-by-gene}
data %>%
	filter(Family_short == "rrn") %>%
	group_by(Gene) %>%
	summarize(Count = n(), Size = sum(Size)) %>%
	kable
```

# Tabulate transfer RNA (tRNA) genes
```{r tabulate-trna-by-gene}
data %>%
	filter(Family_short == "trn") %>%
	group_by(Gene) %>%
	summarize(Count = n(), Size = sum(Size)) %>%
	mutate(Index = row_number()) %>%
	select(Index, everything()) %>%
	kable
```

# Tabulate protein-coding genes (mRNA) by family
```{r tabulate-mrna-by-family}
tab <- data_grouped %>%
	filter(!Family_short %in% c("orf", "rrn", "trn")) %>%
	set_rownames(paste(.$Family_short, .$Family)) %>%
	ungroup %>%
	select(Count, Size)
tab['Sum',] <- colSums(tab)
kable(tab)
```

# Tabulate protein-coding genes (mRNA) by gene
```{r tabulate-mrna-by-gene}
data %>%
	filter(!Family_short %in% c("orf", "rrn", "trn")) %>%
	summarize(Count = n(), Distinct = n_distinct(Gene), Size = sum(Size)) %>%
	kable

data %>%
	filter(!Family_short %in% c("orf", "rrn", "trn")) %>%
	group_by(Gene) %>%
	summarize(Count = n(), Size = sum(Size)) %>%
	kable
```

# Tabulate open reading frames (ORF)
```{r tabulate-orf}
data_grouped %>% filter(Family_short == "orf") %>% kable
```

# Tabulate open reading frames (ORF) at least 300 bp
```{r tabulate-orf-300bp}
data %>%
	filter(Family_short == "orf" & Size >= 300) %>%
	group_by(Type, Family) %>%
	summarise(
		Count = n(),
		Size = sum(Size)) %>%
	right_join(gene_types, ., by = "Family") %>%
	kable
```

# List other genes
```{r list-other-genes}
data %>% filter(Family == "other genes") %>% kable
```

# Tabulate introns
```{r tabulate-introns}
introns_gff <- gff_all[gff_all$type == "intron",]

introns <- data.frame(
		getGffAttribute(introns_gff, 'Parent'),
		Size = size(introns_gff),
		stringsAsFactors = FALSE) %>%
	left_join(data, by = c("Parent" = "ID")) %>%
	rename(Intron_size = Size.x, Gene_size = Size.y) %>%
	select(Type, Name, Intron_size, Gene_size) %>%
	arrange(Type, Name) %>%
	mutate(
		Name = sub('[|].*', '', Name),
		Intron_size_mod3 = as.character(Intron_size %% 3))

introns %>%
	group_by(Type) %>%
	summarize(Count = n(), Distinct = n_distinct(Name), Size = sum(Intron_size))

introns %>%
	group_by(Type, Name) %>%
	summarize(Count = n(), Size = sum(Intron_size)) %>%
	kable
```

# Tabulate introns at least 300 bp
```{r tabulate-introns-large}
introns_large <- introns %>% filter(Intron_size >= 300)

introns_large %>%
	group_by(Type) %>%
	summarize(Count = n(), Distinct = n_distinct(Name), Size = sum(Intron_size))

introns_large %>%
	group_by(Type, Name) %>%
	summarize(Count = n(), Size = sum(Intron_size)) %>%
  ungroup()
```

# Histogram of intron size
```{r intron-size-histogram}
ggplot(introns) + theme_bw() +
	aes(x = Intron_size, fill = Intron_size_mod3, group = Intron_size_mod3) +
	geom_histogram(binwidth = 100, boundary = 0) +
	scale_x_continuous(name = "Size (bp)", breaks = seq(0, 5000, 500)) +
	scale_y_continuous(name = "Count", breaks = seq(0, 20, 1))
```

# Histogram of gene size (≤ 1000 bp)
```{r gene-size-histogram}
ggplot(data) + theme_bw() +
	aes(x = Size) +
	geom_histogram(binwidth = 50, boundary = 0) +
	scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 100)) +
	xlab("Size (bp)")
```

# Gene size by family
```{r gene-size-by-family-strip}
data %>%
	filter(Family_short != "orf") %>%
ggplot() + theme_bw() +
	aes(x = Size / 1e3, y = Family, colour = Gene) +
	geom_point() +
	scale_colour_manual(values = rep(brewer.pal(12, "Set3"), length.out = 99)) +
	xlab('Size (kbp)')
```

# Total gene size by family
```{r stripplot-total-gene-size-by-family, fig.height = 9}
data %>%
	filter(Family_short != "orf") %>%
	group_by(Family, Gene) %>%
	summarize(Size = sum(Size)) %>%
ggplot() + theme_bw() +
	aes(x = Size / 1e3, y = Family, colour = Gene) +
	geom_point() +
	scale_colour_manual(values = rep(brewer.pal(12, "Set3"), length.out = 99)) +
	xlab('Size (kbp)')
```

# Gene counts by family
```{r gene-counts-by-family}
data %>%
	filter(Family_short != "orf") %>%
ggplot() + theme_bw() +
	aes(x = Family, fill = Gene) +
	geom_bar(colour = "black") +
	coord_flip() +
	scale_fill_manual(values = rep(brewer.pal(12, "Set3"), length.out = 99))
```

# Gene size by family
```{r gene-size-by-family, fig.width = 7, fig.height = 7, dpi = 300}
data_barchart <- data %>%
	filter(Family_short != "orf") %>%
	mutate(Family = str_wrap(paste(Family_short, Family, sep = ": "), 30),
		Gene = Gene %>% sub("trn.*", "trn*", .) %>%
			sub("ymf.*", "ymf*", .)) %>%
	arrange(Gene)
ggplot(data_barchart) +
	theme_bw() +
	theme(legend.key = element_rect(colour = "black"),
		legend.text = element_text(size = 10)) + 
	aes(y = Size / 1e3, x = Family, fill = Gene) +
	geom_bar(stat = "identity", color = "black") +
	coord_flip() +
	scale_fill_manual(values = rep(brewer.pal(12, "Set3"), length.out = 99)) +
	guides(fill = guide_legend(keyheight = 0.6, override.aes = list(colour = NULL))) +
	xlim(data_barchart$Family %>% unique %>% sort(decreasing = TRUE)) +
	ylab('Size (kbp)')
```

# Gene size by scaffold
```{r gene-size-by-scaffold, fig.width = 8, fig.height = 9, dpi = 300}
data_scaffolds <- data %>%
	filter(Family_short != "orf") %>%
	mutate(Family = factor(Family_short, levels = gene_types$Family_short, labels = gene_types$Family_both))

Family_count <- length(unique(data_scaffolds$Family))

ggplot(data_scaffolds) +
	aes(y = Size, x = Scaffold, fill = Family,
		label = ifelse(Size >= 1000, Gene, "")) +
	geom_bar(stat = "identity", color = "black") +
	coord_flip() +
	geom_text(position = "stack", hjust = 1.05) +
	scale_fill_manual("", breaks = gene_types$Family_both, values = gene_types$Colour) +
	scale_x_discrete(limits = rev(unique(data_scaffolds$Scaffold))) +
	scale_y_continuous(labels = unit_format(unit = "kbp", scale = 1e-3)) +
	theme_bw() +
    theme(
        legend.key = element_rect(colour = "black"),
	    legend.justification = c(1, 0),
        legend.position = c(1, 0)) +
	guides(fill = guide_legend(ncol = 2, override.aes = list(colour = NULL)))
```


# Function to plot gene order and size
```{r gene-order-and-size-function}
data_genes <- data %>%
	filter(Family_short != "orf") %>%
	mutate(
		Family = factor(Family_short, levels = gene_types$Family_short, labels = gene_types$Family_both),
		Gene = sub("ymf.*", "ymf*", Gene),
		Size = pmax(Size, 200),
		TextSize = pmax(Size, 2000) + 100) %>%
	replace_na(list(Family = "oth: other genes")) %>%
	left_join(scaffold_sizes, by = "Scaffold") %>%
	arrange(desc(ScaffoldSize)) %>%
	mutate(ScaffoldNameAndSize = factor(ScaffoldNameAndSize, unique(ScaffoldNameAndSize))) %>%
	group_by(Scaffold) %>%
	mutate(x = cumsum(TextSize) - TextSize / 2) %>%
	ungroup()

plot_gene_order_and_size <- function(data)
ggplot(data) +
	theme_bw(base_size = 12) +
	theme(axis.ticks.y = element_blank(),
		axis.title.x = element_text(hjust = 0.04),
		axis.title.y = element_text(hjust = 1),
		legend.justification = c(1, 0),
		legend.position = c(1, 0),
		panel.border = element_rect(colour = NA),
		panel.grid = element_blank(),
		strip.text = element_text(hjust = 0),
		strip.background = element_rect(colour = NA, fill = NA)) + 
	aes(x = x / 1e3,
		y = Strand,
		width = Size / 1e3,
		fill = Family,
		label = Gene) +
	facet_wrap(~ ScaffoldNameAndSize, ncol = 1) +
	geom_tile() +
	geom_text(size = 1.5, vjust = 2.5) +
	geom_segment(aes(xend = (x + Size/2 + 1000)/1e3), x = 0, y = 2.5, yend = 2.5, size = 0.2) +
	guides(fill = guide_legend(ncol = 2, override.aes = list(colour = NULL))) +
	scale_fill_manual("", values = gene_types %$% setNames(Colour, Family_both)) +
	scale_x_continuous("Gene size (kbp)", breaks = 0:5) +
	ylab("") +
	ylim("", "-", "+")
```

# Gene order and size
```{r gene-order-and-size, fig.width = 7, fig.height = 8, dpi = 300}
data_genes %>% plot_gene_order_and_size()
```

# Gene position
```{r gene-position, fig.width = 7, fig.height = 9, dpi = 300}
data_position <- data %>%
	filter(Family_short != "orf") %>%
	mutate(
		Gene = Gene %>% sub("ccm.*", "ccm*", .) %>% sub("ymf.*", "ymf*", .),
		StrandInt = 3 - 2 * as.integer(Strand),
		Size = pmax(Size, 2000)) %>%
	group_by(Scaffold) %>%
	mutate(Rank = rank(Position) - 1) %>%
	ungroup() %>%
	mutate(x = Rank / max(Rank) * max(Position)) %>%
	left_join(scaffold_sizes, by = "Scaffold")

ggplot(data_position) +
	theme_bw(base_size = 12) +
	theme(axis.ticks.y = element_blank(),
		axis.title.x = element_text(hjust = 0.04),
		axis.title.y = element_text(hjust = 1),
		legend.key = element_rect(colour = "black"),
		legend.justification = c(1, 0),
		legend.position = c(1, 0),
		panel.border = element_rect(colour = NA),
		panel.grid = element_blank(),
		strip.text = element_text(hjust = 0),
		strip.background = element_rect(colour = NA, fill = NA)) + 
	facet_wrap(~ ScaffoldNameAndSize, ncol = 1) +
	geom_tile(aes(
			x = Position / 1e3,
			y = 0.5 * StrandInt,
			width = Size / 1e3,
			fill = Family),
		height = 1.5) +
	geom_segment(aes(
			x = x / 1e3,
			xend = Position / 1e3,
			y = 3 * StrandInt,
			yend = 1.5 * StrandInt),
		colour = "black",
		size = 0.1) +
	geom_text(aes(
			x = x / 1e3,
			y = 5 * StrandInt,
			label = Gene),
		size = 1.75) +
	geom_segment(data = scaffold_sizes %>%
			filter(Scaffold %in% data_position$Scaffold),
		aes(
			x = 0,
			xend = ScaffoldSize / 1e3,
			y = 0,
			yend = 0),
		size = 0.1) +
	guides(fill = guide_legend(override.aes = list(colour = NULL))) +
	scale_fill_manual("", values = gene_types$Colour) +
	scale_y_continuous(limits = c(-6.5, 6),
		breaks = c(-3, 3),
		labels = c("-", "+")) +
	xlab("Position (kbp)") +
	ylab("")
```

# Store the data grouped by family
```{r write-data-by-family}
write_tsv(data_grouped, "psitchensismt_18.gff.family.tsv")
```
